---
title: "Loan Approval Prediction"
author: "Harsh Mishrikotkar"
date: "2025-05-20"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, error = FALSE)
```

### Importing libraries and data

```{r import libraries and data}

# Installing Libraries
library(tidyverse)
library(gridExtra)
library(grid)
library(corrplot)
library(caTools) 
library(class)
library(caret)

# Installing Data set
loanData <- read_csv("loan_approval_dataset.csv")

```
###Variables

* **no_of_dependents:** How many dependents rely on the loan applicant
* **Education:** Weather the loan applicant has or has not graduated
* **self_employed:** Weather the loan applicant is or is not self employed
* **Income_annum:** Annual income of the applicant
* **loan_amount:** Amount of loan application
* **loan_term:** Loan term in years
* **cibil_score:** Credit score
* **residential_assets_value:** Value of residential assets
* **commercial_assets_value:** Value of commercial assets
* **luxury_assets_value:** Value of luxury assets
* **bank_asset_value:** Value of bank assets
* **loan_status:** Weather the loan was approved or rejected

### Data Cleaning

```{r Data Cleaning}

loanData <- loanData|>
  select(-loan_id)

summary(loanData)
colSums(is.na(loanData))

```

### Graphs of Variables

```{r Graph of variables, fig.width = 10, fig.height = 30}

graphMaker <- function(data) {
  
  numeric_cols <- names(data)[sapply(data, is.numeric)]
  categorical_cols <- names(data)[sapply(data, function(x) is.factor(x) | is.character(x))]
  
  categorical_cols <- categorical_cols[categorical_cols != "loan_status"]
  
  if ("loan_status" %in% names(data)) {
    data$loan_status <- as.factor(data$loan_status)
  } else {
    data$loan_status <- factor("None")
    }
  
  loan_colors <- c("Approved" = "green", "Rejected" = "red")
  
  plot_list <- list()
  plot_count <- 1
  
  for (col in numeric_cols) {
    p <- ggplot(data,
                aes_string(x = col,
                           fill = "loan_status")) +
      geom_histogram(position = "identity",
                     alpha = 0.6,
                     bins = 30,
                     colour = "black") +
      scale_fill_manual(values = loan_colors) +
      labs(title = paste("Histogram of", col),
           x = col,
           y = "Count") +
      theme_minimal() +
      theme(legend.title = element_blank())
    
    plot_list[[plot_count]] <- p
    plot_count <- plot_count + 1
  }
  
  for (col in categorical_cols) {
    
    p <- ggplot(data,
                aes_string(x = col, fill = "loan_status")) +
      geom_bar(position = "dodge",
               colour = "black") +
      scale_fill_manual(values = loan_colors) +
      labs(title = paste("Bar Chart of", col),
           x = col,
           y = "Count") +
      theme_minimal() +
      theme(legend.title = element_blank(),
            axis.text.x = element_text(angle = 45, hjust = 1))
    
    plot_list[[plot_count]] <- p
    plot_count <- plot_count + 1
  }
  
  n_rows <- ceiling(length(plot_list) / 2)
  
  if (length(plot_list) > 0) {
    grid.arrange(grobs = plot_list, ncol = 2)
  } else {
    cat("No plots were created. Check your data.")
    }
}

graphMaker(loanData)


```

With the histograms and bar charts above: Rejected loan applications occurred almost guaranteed when the credit score was below approximately 500. Otherwise all the variables have a similar spread between Rejected and Approved


```{r Coorelation Plot, fig.width = 15, fig.height = 15}

cols = c(4, 5, 7:11)
pairedData <- loanData
pairs(pairedData[, cols], panel=function(x,y){
  colors <- ifelse(loanData$loan_status == "Approved", "green",
                   ifelse(loanData$loan_status == "Rejected", "red", "blue"))
  points(x, y, col = colors)
})

```

Given the pair plots, we have limits on all variables, with none, other than the Credit Score having any distinction for any patterns.

```{r KNN model}

set.seed(1763)

loanApprovalModel <- function(data, test_size = 0.3, k = NULL) {
  
  data$education <- as.factor(data$education)
  data$self_employed <- as.factor(data$self_employed)
  data$loan_status <- as.factor(data$loan_status)
  
  numeric_cols <- c("no_of_dependents",
                    "income_annum",
                    "loan_amount",
                    "loan_term",
                    "cibil_score",
                    "residential_assets_value",
                    "commercial_assets_value",
                    "luxury_assets_value",
                    "bank_asset_value")
  data_processed <- data
  
  data_processed[numeric_cols] <- lapply(data_processed[numeric_cols], function(x) {
    (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
  })
  
  data_processed$education_graduate <- ifelse(data_processed$education == "Graduate", 1, 0)
  data_processed$self_employed_yes <- ifelse(data_processed$self_employed == "Yes", 1, 0)
  
  data_processed$education <- NULL
  data_processed$self_employed <- NULL
  
  sample_index <- createDataPartition(data_processed$loan_status, p = 1-test_size, list = FALSE)
  train_data <- data_processed[sample_index, ]
  test_data <- data_processed[-sample_index, ]
  
  train_features <- train_data[, !(names(train_data) %in% c("loan_status"))]
  train_target <- train_data$loan_status
  
  test_features <- test_data[, !(names(test_data) %in% c("loan_status"))]
  test_target <- test_data$loan_status
  
  if(is.null(k)) {
    k_values <- seq(1, min(29, nrow(train_data)-1), by = 2)
    accuracy_values <- numeric(length(k_values))
    
    for(i in seq_along(k_values)) {
      knn_pred <- knn(train = train_features,
                      test = test_features,
                      cl = train_target,
                      k = k_values[i])
      
      accuracy_values[i] <- mean(knn_pred == test_target)
    }
    
    k <- k_values[which.max(accuracy_values)]
  }
  
  final_pred <- knn(train = train_features,
                    test = test_features,
                    cl = train_target,
                    k = k,
                    prob = TRUE)
  
  conf_matrix <- table(Predicted = final_pred, Actual = test_target)
  accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
  
  return(list(train_features = train_features,
              train_target = train_target,
              test_features = test_features,
              test_target = test_target,
              optimal_k = k,
              predictions = final_pred,
              confusion_matrix = conf_matrix,
              accuracy = accuracy,
              model = function(new_data) {
                return(predict_loan_status(new_data, train_features, train_target, k))
              }))
}

model_results <- loanApprovalModel(loanData)

print(model_results$confusion_matrix)

print(paste("Accuracy:", model_results$accuracy))

print(paste("Optimal k:", model_results$optimal_k))

```

The KNN model scaled to the normal values has an accuracy of `r round(model_results$accuracy, 4)` or `r round(model_results$accuracy, 4)*100`%.

This model is accurate as it has an accuracy above `r round(model_results$accuracy, 4)*100`%. This is enough to belive the model to be accurate.